---
title: "batch effects"
author: "Jesse Cohen"
date: "`r Sys.Date()`"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(ROCR)
library(caret)
library(devtools)
install_github("jfortin1/neuroCombat_Rpackage")
library(neuroCombat)
devtools::install_github("andy1764/CovBat_Harmonization/R")
library(CovBat)
library(magrittr)
library(janitor)
library(gt)
library(plotly)

rm(list=ls())

```


```{r Data Import Template, echo=FALSE, results='hide'}
# input
mri_bids_data_t1 <- read_rds(here("objects/mri_bids_data_t1.RDS"))
ashs_clean <- read_rds("objects/ashs_clean.RDS")

# Initialise the dataset as per the template.
dsname <- "mri_bids_data_t1"
ds     <- get(dsname)

ds %>% sample_frac()

glimpse(ds)
```


```{r Normalize variable names automatically, results='hide', echo=FALSE}
# Normalise the variable names.

ds %<>% clean_names(numerals="right")

# Confirm the results are as expected.

names(ds)

```

```{r split dicom_series_instance}
ds %<>%
  mutate(scanner_id = str_match(dicom_series_instance_uid, "((\\d+\\.){9})")[,3]) %>% 
  select(inddid, flywheel_session_label, dicom_series_instance_uid, scanner_id,
         flywheel_acquisition_label, everything())

# To get rid of the NAs created:
# ds$scanner_id[grepl(pattern = "(1.2.826.0.1.3680043.8.498.)*", x = ds$dicom_series_instance_uid)]

# See how many new NAs introduced 
ds$dicom_series_instance_uid %>% is.na() %>% sum()
ds$scanner_id %>% is.na() %>% which() -> na_rows

ds[na_rows,]

```


```{r , results='hide', echo=FALSE}
# Note the available variables.

vars <- names(ds)

# Note the identifiers.

id <- c("inddid", "flywheel_session_label", "bids_filename")

# Initialise ignored variables: identifiers.

ignore <- c(id)

```

```{r , results='hide', echo=FALSE}
#Once we have identified all of the variables to ignore we remove them from our list of variables to use.

# Check the number of variables currently.

length(vars)

# Remove the variables to ignore.

vars <- setdiff(vars, ignore)

# Confirm they are now ignored.

length(vars)
```

```{r Character variables, results='hide', echo=FALSE }
# Identify the character variables by index.

ds[vars] %>%
  sapply(is.character) %>%
  which()  ->
chari

# Identify the character variables by name.

ds[vars] %>% 
  names() %>% 
  '['(chari) ->
charc

# Observe the unique levels.

ds[charc] %>% select(dicom_station_name, scanner_id) %>% sapply(unique)

# # Which columns have > 5 unique levels?
# charc_long <- 
#   ds[charc][sapply(ds[charc], function(x) length(unique(x)) > 5)] %>% 
#   colnames()
# 
# # Ignore character variables with too many levels
# 
# ignore <- union(ignore, charc_long)

```

```{r Numeric Variables ,results='hide', echo=FALSE}
# ds[vars] %>%
#   sapply(is.numeric) %>%
#   which() %>%
#   names ->
# numi
# 
# ds[numi] %>% 
#   summary()
```

```{r remove columns from ds before joining, results='hide', echo=FALSE}
ds %<>% select(inddid, flywheel_session_label, scanner_id, contains("dicom")) %>% distinct()
```

```{r join ashs with bids}
ds1 <-
  ashs_clean %>% 
  left_join(ds) %>% 
  distinct() 


ds1$dicom_series_instance_uid %>% is.na() %>% sum()
ds1$scanner_id %>% is.na() %>% which() -> na_rows

which(grepl("3680043.8.498", as.character(ds1$dicom_series_instance_uid))) -> indices

setdiff(na_rows, indices)

# # Convert bids/dicom numerical variables to factors
# ds1[numi] <- round(ds1[numi])
# 
# ds1[numi] <- sapply(ds1[numi], as.factor)
# 
# # filter out duplicates - need to check to make sure this isn't causing the problem 
# ds1  %<>% filter(dicom_slice_thickness == 1 &
#                   dicom_spacing_between_slices !=8) %>% 
#   distinct()


```

```{r simplify factors for batch correction}
# character variables
fct_count(ds1$dicom_station_name) %>% arrange(desc(n))
ds1$dicom_station_name <-
ds1$dicom_station_name %>%
  replace_na("other") %>% 
  as.factor()
  # fct_collapse(
  #   hup6 = c("hup6", "HUP FNDBA MR2", "MRC35154", "MRC20523"),
  #   dr1 = c("dr1", "HUP DEVGR MR1", "c93verio", "hup devon mrd1",
  #           "MRC21236", "other")
  # )
  
# I think the NAs introduced in scanner_id are from PDC since 
# they're almost all PD/PDD subjects (except 1 control), 
# and the dicom_series_instance_iud ends in 
# 3680043.8.498
fct_count(ds1$scanner_id) %>% arrange(desc(n))
ds1$scanner_id <- 
  ds1$scanner_id %>% 
    replace_na("other") %>% 
    as.factor()

# check the station_name associated with each number; for HUP6 (35069), 
# split factor into 
# before May 1, 2017 and after since taht is when the scanner was upgraded 
# from trimtrio to prisma

ds2 <- ds1 %>% 
  mutate(recent_scan = session_date > as.Date("2017-05-01"),
  batch = paste(dicom_station_name, scanner_id, recent_scan, sep = "_")) 
  
ds2 %>% 
group_by(batch) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))


# Remove any remaining duplicates after collapsing factors
fct_count(ds2$batch) %>% arrange(desc(n))
ds2$batch <-
  fct_collapse(
  ds2$batch,
  hup_35154_pre_2017 = c("dr1_35154._FALSE", 
                         "HUP DEVGR MR1_35154._FALSE",
                        "hup devon mrd1_35154._FALSE"),
  hup_167024 = c("HUP FNDBA MR2_167024._TRUE"),
  hup_35154_post_2017 = c("HUP DEVGR MR1_35154._TRUE",
                          "MRC35154_35154._TRUE"),
  hup_35069 = c("hup6_35069._FALSE", 
                "HUP FNDBA MR2_35069._FALSE"),
  hup_20523 = c("MRC20523_20523._FALSE", 
              "MRC20523_other_FALSE"),
  other = c("NA_other_FALSE", 
                  "c93verio_40171._FALSE", 
                  "MRC21236_21236._FALSE")
)


ds2 %<>% select(-contains("dicom")) %>% distinct()
ds2 %>% filter(duplicated(inddid))

```

```{r collapse factors, results='hide', echo=FALSE}
# # Slice thickness
# fct_count(ds1$dicom_slice_thickness)
# ds1$dicom_slice_thickness <- fct_collapse(
#   ds1$dicom_slice_thickness, 
#   `1` = c("1", "2"),
#   `4` = c("4", "5"))
# fct_count(ds1$dicom_slice_thickness)
# 
# # Pixel spacing
# fct_count(ds1$dicom_pixel_spacing_x)
# ds1$dicom_pixel_spacing_x <- fct_collapse(
#   ds1$dicom_pixel_spacing_x, 
#   `1` = c("0", "1"))
# fct_count(ds1$dicom_pixel_spacing_x)
# 
# # Field strength
# fct_count(ds1$dicom_magnetic_field_strength)
# ds1$dicom_magnetic_field_strength <- fct_collapse(
#   ds1$dicom_magnetic_field_strength, 
#   `1.5` = c("1", "15000"),
#   `3` = c("2", "3"))
# fct_count(ds1$dicom_magnetic_field_strength)
# 
# # Repetition 
# fct_count(ds1$dicom_repetition_time)
# ds1$dicom_repetition_time <- fct_collapse(
#   ds1$dicom_repetition_time, 
#   `10` = c("10", "11", "440", "700"),
#   `1620` = c("1140", "1620", "1800", "1900", "2300", "2400", "3000"))
# fct_count(ds1$dicom_repetition_time)
# 
# # Spacing between slices
# fct_count(ds1$dicom_spacing_between_slices)
# ds1$dicom_spacing_between_slices <- fct_collapse(
#   ds1$dicom_spacing_between_slices, 
#   `8` = c("5", "6", "8"))
# fct_count(ds1$dicom_spacing_between_slices)

```



I need to see which files ashs was run with to make sure the other slice thickness with the same session label wasn't used for ashs

```{r Data Exploration, results = TRUE}
#A random sample of the dataset:

# ds1 %>% sample_frac()
# 
# ds1 %>% count(flywheel_session_label) %>% dplyr::arrange(desc(n))
```


```{r script from Taki}

####################################
### TRY WITH RAW DATA - PREDICTING STATION NAME
mtl_nums <- 
  ds2 %>% 
  select((contains("left_") | 
            contains("right_")) &
           -contains("sul"))

y_raw <- mtl_nums
y_raw$y <- as.factor(ds2$batch=="hup_35069")
set.seed(9458)
inTraining <- createDataPartition(y_raw$y, p = .5, list = FALSE)
training <- y_raw[ inTraining,]
testing  <- y_raw[-inTraining,]
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10, repeats = 1)
Fit1 <- train(y ~ .,
              method = "rf", data=training,
              trControl = fitControl)

test_prob <- predict(Fit1, testing,type='prob')[,2]
pred <- prediction(test_prob, testing$y)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
auc1<-performance(pred, measure = "auc")@y.values[[1]]
print(auc1)
plot(perf)

####################################
### TRY WITH RAW DATA - PREDICTING DX
y_raw$y<-factor(ds2$diagnosis=='Normal')
set.seed(9458)
inTraining <- createDataPartition(y_raw$y, p = .5, list = FALSE)
training <- y_raw[ inTraining,]
testing  <- y_raw[-inTraining,]
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10,repeats = 1)
Fit1 <- train(y ~ .,
              method = "rf", data=training,
              trControl = fitControl)

test_prob <- predict(Fit1, testing,type='prob')[,2]
pred <- prediction(test_prob, testing$y)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
auc1<-performance(pred, measure = "auc")@y.values[[1]]
auc1_diagnosis_raw <- print(auc1)
roc_diagnosis_raw <- plot(perf, col=rainbow(10)) 
```
## Batch correction successful
```{r}
### TRY WITH COMBAT DATA
age <- ds2$age_at_mri
disease <- as.factor(ds2$diagnosis != "Normal")
sex <- ds2$sex
education <- ds2$education

y_combat <- data.frame(t(neuroCombat(t(mtl_nums), 
                                     bat = as.numeric(factor(ds2$batch)), 
                                     mod = model.matrix(~ age + sex + disease +
                                                          education,
                                                        data = ds2))$dat.combat))

####################################
### TRY WITH COMBAT DATA - PREDICTING DICOM STATION NAME
y_combat$y<-factor(ds2$batch=='hup_35069')
set.seed(9458)
inTraining <- createDataPartition(y_combat$y, p = .5, list = FALSE)
training <- y_combat[ inTraining,]
testing  <- y_combat[-inTraining,]
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10, repeats = 1)
Fit1 <- train(y ~ .,
              method = "rf", data=training,
              trControl = fitControl)
test_prob <- predict(Fit1, testing,type='prob')[,2]
pred <- prediction(test_prob, testing$y)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
auc1<-performance(pred, measure = "auc")@y.values[[1]]
print(auc1)
plot(perf, col=rainbow(10))

####################################
### TRY WITH COMBAT DATA - PREDICTING DX
y_combat$y<-factor(ds2$diagnosis=='Normal')
set.seed(9458)
inTraining <- createDataPartition(y_combat$y, p = .5, list = FALSE)
training <- y_combat[ inTraining,]
testing  <- y_combat[-inTraining,]
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10, repeats = 1)
Fit1 <- train(y ~ .,
              method = "rf", data=training,
              trControl = fitControl)
test_prob <- predict(Fit1, testing,type='prob')[,2]
pred <- prediction(test_prob, testing$y)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
auc1<-performance(pred, measure = "auc")@y.values[[1]]
print(auc1)
plot(perf, col=rainbow(10))

```

### TRY WITH COMBAT DATA- FORGET DX ADJUSTMENT
This should make it harder to accurately predict the station name, 
but also will washout disease effects making it harder to predict disease
```{r}
y_combat <- data.frame(t(neuroCombat(t(mtl_nums), bat = as.numeric(factor(ds2$batch)))$dat.combat))
```

### TRY WITH COMBAT DATA - PREDICTING DICOM STATION NAME
```{r}
y_combat$y<-factor(ds2$batch=='hup_35069')
set.seed(9458)
inTraining <- createDataPartition(y_combat$y, p = .5, list = FALSE)
training <- y_combat[ inTraining,]
testing  <- y_combat[-inTraining,]
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10, repeats = 1)
Fit1 <- train(y ~ .,
              method = "rf", data=training,
              trControl = fitControl)
test_prob <- predict(Fit1, testing,type='prob')[,2]
pred <- prediction(test_prob, testing$y)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
auc1<-performance(pred, measure = "auc")@y.values[[1]]
print(auc1)
plot(perf, col=rainbow(10))

####################################
### TRY WITH COMBAT DATA - PREDICTING DX
y_combat$y<-factor(ds2$diagnosis=='Normal')
set.seed(9458)
inTraining <- createDataPartition(y_combat$y, p = .5, list = FALSE)
training <- y_combat[ inTraining,]
testing  <- y_combat[-inTraining,]
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10, repeats = 1)
Fit1 <- train(y ~ .,
              method = "rf", data=training,
              trControl = fitControl)
test_prob <- predict(Fit1, testing,type='prob')[,2]
pred <- prediction(test_prob, testing$y)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
auc1<-performance(pred, measure = "auc")@y.values[[1]]
print(auc1)
plot(perf, col=rainbow(10))

```

```{r}
### TRY WITH COVBAT DATA
y_covbat <- data.frame(t(covbat(t(mtl_nums), bat = as.numeric(factor(ds2$batch)),
                                mod = model.matrix(~ age + sex + disease + education,
                                                   data = ds2))$dat.covbat))

####################################
### TRY WITH COVBAT DATA - PREDICTING DICOM STATION NAME
y_covbat$y<-factor(ds2$batch=='hup_35069')
set.seed(9458)
inTraining <- createDataPartition(y_covbat$y, p = .5, list = FALSE)
training <- y_covbat[ inTraining,]
testing  <- y_covbat[-inTraining,]
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10, repeats = 1)
Fit1 <- train(y ~ .,
              method = "rf", data=training,
              trControl = fitControl)
test_prob <- predict(Fit1, testing,type='prob')[,2]
pred <- prediction(test_prob, testing$y)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
auc1<-performance(pred, measure = "auc")@y.values[[1]]
print(auc1)
plot(perf)

####################################
### TRY WITH COVBAT DATA - PREDICTING DX
y_covbat$y<-factor(ds2$diagnosis=='Normal')
set.seed(9458)
inTraining <- createDataPartition(y_covbat$y, p = .5, list = FALSE)
training <- y_covbat[ inTraining,]
testing  <- y_covbat[-inTraining,]
fitControl <- trainControl(## 10-fold CV
  method = "repeatedcv",
  number = 10, repeats = 1)
Fit1 <- train(y ~ .,
              method = "rf", data=training,
              trControl = fitControl)
test_prob <- predict(Fit1, testing,type='prob')[,2]
pred <- prediction(test_prob, testing$y)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
auc1<-performance(pred, measure = "auc")@y.values[[1]]
print(auc1)
plot(perf)

```

<!-- # Combat -->

<!-- ## **1. Multi-Site Harmonization** -->

<!-- ComBat estimates scanner-specific location and scale parameters, for each feature separately, and pools information across features using empirical Bayes to improve the estimation of those parameters for small sample size studies. -->

<!-- ### **1.1 Full ComBat with empirical Bayes** -->

<!-- The `neuroCombat` function is the main function. It requires two mandatory arguments: - a data matrix (p x n) `dat` for which the p rows are features, and the n columns are participants. - a numeric or character vector `batch` of length n indicating the site/scanner/study id. -->

<!-- For illustration purpose, let's simulate an imaging dataset with n=10 participants, acquired on 2 scanners, with 5 participants each, with p=10000 voxels per scan. -->

<!-- ```{r} -->
<!-- library(neuroCombat) -->
<!-- p=10000 -->
<!-- n=10 -->
<!-- batch = c(1,1,1,1,1,2,2,2,2,2) #Batch variable for the scanner id -->
<!-- dat = matrix(runif(p*n), p, n) #Random Data matrix -->

<!-- # Extract batch variable -->
<!-- batch <- as.factor(ds1$dicom_station_name) -->

<!-- # Remove ids from ds1  -->

<!-- ignore <- c("inddid",  -->
<!--             "diagnosis", -->
<!--             "ad_present", -->
<!--             "session_date", -->
<!--             "flywheel_session_label", -->
<!--             "yob", -->
<!--             "race", -->
<!--             "session_year", -->
<!--             "dicom_station_name", -->
<!--             "sex", -->
<!--             "education", -->
<!--             "age_at_mri") -->


<!-- dat <- ds1 %>% select(-c(any_of(ignore))) %>%  -->
<!--   as.matrix() %>% t() -->

<!-- ``` -->

<!-- We use the function `neuroCombat` to harmonize the data across the 2 scanners: -->

<!-- ```{r} -->
<!-- data.harmonized <- neuroCombat(dat=dat, batch=batch) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- # Combine the harmonized data with the site data, covariates, IV  -->
<!-- output <- cbind(t(data.harmonized$dat.combat), ds1[ignore]) %>%  -->
<!--   select(ignore, everything()) %>%  -->
<!--   as.tibble() -->

<!-- ``` -->

<!-- By default, this uses parametric adjustments. To following command must be used for non-parametric adjustments: -->

<!-- ```{r} -->
<!-- data.harmonized <- neuroCombat(dat=dat, batch=batch, parametric=FALSE) -->
<!-- ``` -->

<!-- The harmonized matrix is stored in `data.harmonized$dat.combat`. The `data.harmonized` object also contains the different parameters estimated by ComBat: - `gamma.hat` and `delta.hat`: Estimated location and shift (L/S) parameters before empirical Bayes. - `gamma.star` and `delta.star`: Empirical Bayes estimated L/S parameters. - `gamma.bar`, `t2`, `a.prior` and `b.prior`: esimated prior distributions parameters. -->

<!-- `neuroCombat` also accepts an optional argument, `mod`, which is a matrix containing biological covariates, including the outcome of interest. This is recommended to ensure that biological variability is preserved in the harmonization process. For instance, for a study with age and disease covariates, -->

<!-- ```{r} -->
<!-- # age <- c(82,70,68,66,80,69,72,76,74,80) # Continuous variable -->
<!-- # disease <- as.factor(c(1,2,1,2,1,2,1,2,1,2)) # Categorical variable -->

<!-- age <- ds1$age_at_mri -->
<!-- disease <- as.factor(ds1$diagnosis != "Normal") -->
<!-- sex <- ds1$sex -->
<!-- education <- ds1$education -->
<!-- ``` -->

<!-- we first create a model matrix for these two biological covariates using the `model.matrix` function: -->

<!-- ```{r} -->
<!-- # mod <- model.matrix(~age+disease) -->
<!-- # mod  -->

<!-- mod <- model.matrix(~ age + disease + sex + education) -->

<!-- ``` -->

<!-- The matrix `mod` is a n x 3 matrix, containing an intercept, age and a dummy variable for the second level of the disease variable (the first level is taken as the baseline group). Note that including an intercept in the model matrix will not change the results of the algorithm; ComBat automatically removes the intercept from the model matrix when fitting the models. We now harmonize the data: -->

<!-- ```{r} -->
<!-- combat.harmonized <- neuroCombat(dat=dat, batch=batch, mod=mod) -->
<!-- ``` -->
# Unadjusted for ICV, before batch correction
## Boxplots
```{r boxplots}
title <- "Subregion Volumes, Unadjusted (Before Batch Correction)"
input <- ashs_clean
#obtain order of regions by mean volume
df_raw_vol <- input %>% select(inddid, ad_present, contains(c("left_", "right_"))) %>%
  select(-contains(c("misc", "meninges", "sul"))) %>%
  pivot_longer(cols = contains(c("left_", "right_")), names_to = "Region", values_to = "Volume") %>%
  mutate(ad_present = as.factor(ad_present))

df_raw_vol %>%
  group_by(Region) %>%
  summarise(Avg_vol = mean(Volume)) %>% arrange(desc(Avg_vol))

df_raw_vol$Region <- recode_factor(df_raw_vol$Region,
                           left_anterior_hippocampus= "L Ant Hip",
                           right_anterior_hippocampus = "R Ant Hip",
                           left_posterior_hippocampus = "L Post Hip",
                           right_posterior_hippocampus = "R Post Hip",
                           left_br_36 = "L Br36",
                           right_br_36 = "R Br36",
                           left_phc = "L PHC",
                           right_phc = "R PHC",
                           left_br_35 = "L Br35",
                           right_br_35 = "R Br35",
                           left_erc = "L ERC",
                           right_erc = "R ERC",
                           left_icv = "ICV")

df_raw_vol$ad_present <- recode_factor(df_raw_vol$ad_present,
                               Normal = "Control",
                               `FALSE` = "LBD-AD",
                               `TRUE` = "LBD+AD")


plot <-
  ggplot(df_raw_vol, aes(x = ad_present, y = Volume)) +
  geom_boxplot() +
  facet_wrap(.~Region, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = title,
       y = "Volume (cm^3)",
       x = "Case Status") +
  font("y", size = 15) +
  font("x", size = 15) +
  theme(strip.text.x = element_text(size = 7)) +
  geom_jitter(alpha = 0.2, size =1)

plot
```

## Regression - Controls vs Cases
```{r regression}
df <- ashs_clean
title <- c("Unadjusted Vols, Before Batch Correction (Control = 0, Case = 1)")

# Define comparison, i.e. control = 0, case = 1 
df1 <- df %>%  select(-contains(c("misc", "meninges"))) %>%
  mutate(case_status = ifelse(diagnosis == "Normal", 0,
                              ifelse(diagnosis != "Normal", 1,
                                      "NA")))

# Position of variables in data frame to be used in for loop
# region raw volume
vol_start <- grep("left_anterior_hippocampus$", colnames(df1)) #column index to begin taking outcome vars
vol_end <- grep("right_phc$", colnames(df1)) #column index to end
vol_nvar <- vol_end - vol_start + 1

# shift within df
out_shift <- vol_start - 1

#For loop
dat <- df1
out_mod <- vector('list', length(vol_nvar))
i <- vol_start

for (i in vol_start:vol_end){
  raw_volume = colnames(dat)[i]

mod_1 <- lm(get(raw_volume) ~ case_status,
      na.action = na.exclude,
      data=dat)

out_mod[[i-out_shift]] <- mod_1
names(out_mod)[i-out_shift] <- paste0("model_", colnames(dat)[i])
}


library(broom)

out_mod_summary <- summary(out_mod[[1]])
out_mod_summary$coefficients %>% as_tibble()
z <- tidy(mod_1, conf.int =T)[2,]
z %>% mutate(region = names(out_mod[1]))

tidy(out_mod[[1]])

names(out_mod[1])

df <- NULL

for (i in 1:length(out_mod)) {
z <- tidy(out_mod[[i]], conf.int =T)[2,]
z <- z %>% mutate(region = names(out_mod[i]))
df <- rbind(df, z)
}

df1 <- df %>% filter(!str_detect(region, pattern = "_sul|_misc|_meninges")) %>%
  arrange(p.value) %>%
  mutate(region = str_remove(region, "model_"),
         region = str_replace(region, "_", " ")) %>%
  select(Region = "region", `Estimate` = "estimate", p = "p.value")

library(kableExtra)
table <- knitr::kable(df1, digits = 4,
  caption = title)

kable_classic_2(table)

```



## Regression - AD+ vs AD - 
```{r regression}
df <- ashs_clean
title <- c("Not Batch Corrected; not divided by ICV (AD- = 0, AD+ = 1)")

# Define comparison, i.e. control = 0, case = 1 
df1 <- df %>%  select(-contains(c("misc", "meninges"))) %>%
  filter(ad_present != "Control") %>% 
  mutate(case_status = ifelse(ad_present == "LBD-AD", 0,
                              ifelse(ad_present == "LBD+AD", 1,
                                      "NA")))

# Position of variables in data frame to be used in for loop
# region raw volume
vol_start <- grep("left_anterior_hippocampus$", colnames(df1)) #column index to begin taking outcome vars
vol_end <- grep("right_phc$", colnames(df1)) #column index to end
vol_nvar <- vol_end - vol_start + 1

# shift within df
out_shift <- vol_start - 1

#For loop
dat <- df1
out_mod <- vector('list', length(vol_nvar))
i <- vol_start

for (i in vol_start:vol_end){
  raw_volume = colnames(dat)[i]

mod_1 <- lm(get(raw_volume) ~ case_status,
      na.action = na.exclude,
      data=dat)

out_mod[[i-out_shift]] <- mod_1
names(out_mod)[i-out_shift] <- paste0("model_", colnames(dat)[i])
}


library(broom)

out_mod_summary <- summary(out_mod[[1]])
out_mod_summary$coefficients %>% as_tibble()
z <- tidy(mod_1, conf.int =T)[2,]
z %>% mutate(region = names(out_mod[1]))

tidy(out_mod[[1]])

names(out_mod[1])

df <- NULL

for (i in 1:length(out_mod)) {
z <- tidy(out_mod[[i]], conf.int =T)[2,]
z <- z %>% mutate(region = names(out_mod[i]))
df <- rbind(df, z)
}

df1 <- df %>% filter(!str_detect(region, pattern = "_sul|_misc|_meninges")) %>%
  arrange(p.value) %>%
  mutate(region = str_remove(region, "model_"),
         region = str_replace(region, "_", " ")) %>%
  select(Region = "region", `Estimate` = "estimate", p = "p.value")

library(kableExtra)
table <- knitr::kable(df1,
  caption = title)

kable_classic_2(table)

```



# Adjusted for ICV, before batch correction
```{r}
df1 <- ashs_clean
vol_start <- grep("left_anterior_hippocampus$", colnames(df1)) #column index to begin taking outcome vars
vol_end <- grep("left_icv$", colnames(df1)) #column index to end

ashs_clean
ashs_adjusted <- ashs_clean[vol_start:vol_end] %>% transmute(./(left_icv*100)) %>% cbind(ashs_clean[1:(vol_start-1)])

saveRDS(ashs_adjusted, file = "objects/ashs_adjusted_icv.RDS")
```

## Boxplots
```{r boxplots}
title <- "Subregion Volumes, Adjusted (Before Batch Correction)"
input <- ashs_adjusted
#obtain order of regions by mean volume
df_raw_vol <- input %>% select(inddid, ad_present, contains(c("left_", "right_"))) %>%
  select(-contains(c("misc", "meninges", "sul"))) %>%
  pivot_longer(cols = contains(c("left_", "right_")), names_to = "Region", values_to = "Volume") %>%
  mutate(ad_present = as.factor(ad_present))

df_raw_vol %>%
  group_by(Region) %>%
  summarise(Avg_vol = mean(Volume)) %>% arrange(desc(Avg_vol))

df_raw_vol$Region <- recode_factor(df_raw_vol$Region,
                           left_anterior_hippocampus= "L Ant Hip",
                           right_anterior_hippocampus = "R Ant Hip",
                           left_posterior_hippocampus = "L Post Hip",
                           right_posterior_hippocampus = "R Post Hip",
                           left_br_36 = "L Br36",
                           right_br_36 = "R Br36",
                           left_phc = "L PHC",
                           right_phc = "R PHC",
                           left_br_35 = "L Br35",
                           right_br_35 = "R Br35",
                           left_erc = "L ERC",
                           right_erc = "R ERC",
                           left_icv = "ICV")

df_raw_vol$ad_present <- recode_factor(df_raw_vol$ad_present,
                               Normal = "Control",
                               `FALSE` = "LBD-AD",
                               `TRUE` = "LBD+AD")


plot <-
  ggplot(df_raw_vol, aes(x = ad_present, y = Volume)) +
  geom_boxplot() +
  facet_wrap(.~Region, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = title,
       y = "Volume (cm^3)",
       x = "Case Status") +
  font("y", size = 15) +
  font("x", size = 15) +
  theme(strip.text.x = element_text(size = 7)) +
  geom_jitter(alpha = 0.2, size =1)

plot
```


## Regression - Controls vs Cases
```{r regression}
df <- ashs_adjusted
title <- c("Adjusted Vols, Before Batch Correction (Control = 0, Case = 1)")

# Define comparison, i.e. control = 0, case = 1 
df1 <- df %>%  select(-contains(c("misc", "meninges"))) %>%
  mutate(case_status = ifelse(diagnosis == "Normal", 0,
                              ifelse(diagnosis != "Normal", 1,
                                      "NA")))

# Position of variables in data frame to be used in for loop
# region raw volume
vol_start <- grep("left_anterior_hippocampus$", colnames(df1)) #column index to begin taking outcome vars
vol_end <- grep("right_phc$", colnames(df1)) #column index to end
vol_nvar <- vol_end - vol_start + 1

# shift within df
out_shift <- vol_start - 1

#For loop
dat <- df1
out_mod <- vector('list', length(vol_nvar))
i <- vol_start

for (i in vol_start:vol_end){
  raw_volume = colnames(dat)[i]

mod_1 <- lm(get(raw_volume) ~ case_status,
      na.action = na.exclude,
      data=dat)

out_mod[[i-out_shift]] <- mod_1
names(out_mod)[i-out_shift] <- paste0("model_", colnames(dat)[i])
}


library(broom)

out_mod_summary <- summary(out_mod[[1]])
out_mod_summary$coefficients %>% as_tibble()
z <- tidy(mod_1, conf.int =T)[2,]
z %>% mutate(region = names(out_mod[1]))

tidy(out_mod[[1]])

names(out_mod[1])

df <- NULL

for (i in 1:length(out_mod)) {
z <- tidy(out_mod[[i]], conf.int =T)[2,]
z <- z %>% mutate(region = names(out_mod[i]))
df <- rbind(df, z)
}

df1 <- df %>% filter(!str_detect(region, pattern = "_sul|_misc|_meninges")) %>%
  arrange(p.value) %>%
  mutate(region = str_remove(region, "model_"),
         region = str_replace(region, "_", " ")) %>%
  select(Region = "region", `Estimate` = "estimate", p = "p.value", )

library(kableExtra)
table <- knitr::kable(df1, digits = 10,
  caption = title)

kable_classic_2(table)

```



## Regression - AD+ vs AD - 
```{r regression}
df <- ashs_adjusted
title <- c("Not Batch Corrected; divided by ICV (AD- = 0, AD+ = 1)")

# Define comparison, i.e. control = 0, case = 1 
df1 <- df %>%  select(-contains(c("misc", "meninges"))) %>%
  filter(ad_present != "Control") %>% 
  mutate(case_status = ifelse(ad_present == "LBD-AD", 0,
                              ifelse(ad_present == "LBD+AD", 1,
                                      "NA")))

# Position of variables in data frame to be used in for loop
# region raw volume
vol_start <- grep("left_anterior_hippocampus$", colnames(df1)) #column index to begin taking outcome vars
vol_end <- grep("right_phc$", colnames(df1)) #column index to end
vol_nvar <- vol_end - vol_start + 1

# shift within df
out_shift <- vol_start - 1

#For loop
dat <- df1
out_mod <- vector('list', length(vol_nvar))
i <- vol_start

for (i in vol_start:vol_end){
  raw_volume = colnames(dat)[i]

mod_1 <- lm(get(raw_volume) ~ case_status,
      na.action = na.exclude,
      data=dat)

out_mod[[i-out_shift]] <- mod_1
names(out_mod)[i-out_shift] <- paste0("model_", colnames(dat)[i])
}


library(broom)

out_mod_summary <- summary(out_mod[[1]])
out_mod_summary$coefficients %>% as_tibble()
z <- tidy(mod_1, conf.int =T)[2,]
z %>% mutate(region = names(out_mod[1]))

tidy(out_mod[[1]])

names(out_mod[1])

df <- NULL

for (i in 1:length(out_mod)) {
z <- tidy(out_mod[[i]], conf.int =T)[2,]
z <- z %>% mutate(region = names(out_mod[i]))
df <- rbind(df, z)
}

df1 <- df %>% filter(!str_detect(region, pattern = "_sul|_misc|_meninges")) %>%
  arrange(p.value) %>%
  mutate(region = str_remove(region, "model_"),
         region = str_replace(region, "_", " ")) %>%
  select(Region = "region", `Estimate` = "estimate", p = "p.value", 
         `S.E.` = "std.error", p = "p.value", `2.5% CI` = "conf.low", 
         `97.5% CI` = "conf.high")

library(kableExtra)
table <- knitr::kable(df1,
  caption = title)

kable_classic_2(table)

```




# Batch corrected for scanner, unadjusted for ICV
## Boxplots
```{r boxplot}
# check to make sure no shift in rows
plot(y_covbat$left_anterior_hippocampus, ds2$left_anterior_hippocampus) #well correlated

plot(y_covbat$left_anterior_hippocampus, rev(ds2$left_anterior_hippocampus)) #randomly correlated

# bind y_covbat back to ds2
ds_covbat <-
  ds2 %>% 
  select(-(contains("dicom") | contains("left_") |
           contains("right_"))) %>% 
  cbind(y_covbat) %>% 
  select(-y) %>% 
  as.tibble()


#obtain order of regions by mean volume
df_raw_vol <- ds_covbat %>% select(inddid, ad_present, contains(c("left_", "right_"))) %>%
  select(-contains(c("misc", "meninges", "sul"))) %>%
  pivot_longer(cols = contains(c("left_", "right_")), names_to = "Region", values_to = "Volume") %>%
  mutate(ad_present = as.factor(ad_present))

df_raw_vol %>%
  group_by(Region) %>%
  summarise(Avg_vol = mean(Volume)) %>% arrange(desc(Avg_vol))

df_raw_vol$Region <- recode_factor(df_raw_vol$Region,
                           left_anterior_hippocampus= "L Ant Hip",
                           right_anterior_hippocampus = "R Ant Hip",
                           left_posterior_hippocampus = "L Post Hip",
                           right_posterior_hippocampus = "R Post Hip",
                           left_br36 = "L Br36",
                           right_br36 = "R Br36",
                           left_phc = "L PHC",
                           right_phc = "R PHC",
                           left_br35 = "L Br35",
                           right_br35 = "R Br35",
                           left_erc = "L ERC",
                           right_erc = "R ERC")

df_raw_vol$ad_present <- recode_factor(df_raw_vol$ad_present,
                               Normal = "Control",
                               `FALSE` = "LBD-AD",
                               `TRUE` = "LBD+AD")


plot <-
  ggplot(df_raw_vol, aes(x = ad_present, y = Volume)) +
  geom_boxplot() +
  facet_wrap(.~Region, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Subregion Volumes, Unadjusted (After Batch Correction)",
       y = "Volume (cm^3)",
       x = "Case Status") +
  font("y", size = 15) +
  font("x", size = 15) +
  theme(strip.text.x = element_text(size = 7)) +
  geom_jitter(alpha = 0.2, size =1)

plot
```

## Regression - Case vs Control

```{r regression}
df <- ds_covbat
title <- c("Batch Corrected Unadjusted Vols (Control = 0, Case = 1)")

# Define comparison, i.e. control = 0, case = 1 
df1 <- df %>%  select(-contains(c("misc", "meninges"))) %>%
  mutate(case_status = ifelse(diagnosis == "Normal", 0,
                              ifelse(diagnosis != "Normal", 1,
                                      "NA")))

# Position of variables in data frame to be used in for loop
# region raw volume
vol_start <- grep("left_anterior_hippocampus$", colnames(df1)) #column index to begin taking outcome vars
vol_end <- grep("right_phc$", colnames(df1)) #column index to end
vol_nvar <- vol_end - vol_start + 1

# shift within df
out_shift <- vol_start - 1

#For loop
dat <- df1
out_mod <- vector('list', length(vol_nvar))
i <- vol_start

for (i in vol_start:vol_end){
  raw_volume = colnames(dat)[i]

mod_1 <- lm(get(raw_volume) ~ case_status,
      na.action = na.exclude,
      data=dat)

out_mod[[i-out_shift]] <- mod_1
names(out_mod)[i-out_shift] <- paste0("model_", colnames(dat)[i])
}


library(broom)

out_mod_summary <- summary(out_mod[[1]])
out_mod_summary$coefficients %>% as_tibble()
z <- tidy(mod_1, conf.int =T)[2,]
z %>% mutate(region = names(out_mod[1]))

tidy(out_mod[[1]])

names(out_mod[1])

df <- NULL

for (i in 1:length(out_mod)) {
z <- tidy(out_mod[[i]], conf.int =T)[2,]
z <- z %>% mutate(region = names(out_mod[i]))
df <- rbind(df, z)
}

df1 <- df %>% filter(!str_detect(region, pattern = "_sul|_misc|_meninges")) %>%
  arrange(p.value) %>%
  mutate(region = str_remove(region, "model_"),
         region = str_replace(region, "_", " ")) %>%
  select(Region = "region", `Estimate` = "estimate", p = "p.value")

library(kableExtra)
table <- knitr::kable(df1, digits = 4,
  caption = title)

kable_classic_2(table)

```


# Batch Corrected and volume adjusted (divided by ICV)
```{r boxplot}
# check to make sure no shift in rows
plot(y_covbat$left_anterior_hippocampus, ds2$left_anterior_hippocampus) #well correlated

plot(y_covbat$left_anterior_hippocampus, rev(ds2$left_anterior_hippocampus)) #randomly correlated

# bind y_covbat back to ds2
ds_covbat <-
  ds2 %>% 
  select(-(contains("dicom") | contains("left_") |
           contains("right_"))) %>% 
  cbind(y_covbat) %>% 
  select(-y) %>% 
  as.tibble()

ds_covbat_adjusted <- ds_covbat[15:33] %>% transmute(./left_icv) %>% cbind(ds_covbat[1:14])

#obtain order of regions by mean volume
df_raw_vol <- ds_covbat_adjusted %>% select(inddid, ad_present, contains(c("left_", "right_"))) %>%
  select(-contains(c("misc", "meninges", "sul"))) %>%
  pivot_longer(cols = contains(c("left_", "right_")), names_to = "Region", values_to = "Volume") %>%
  mutate(ad_present = as.factor(ad_present))

df_raw_vol %>%
  group_by(Region) %>%
  summarise(Avg_vol = mean(Volume)) %>% arrange(desc(Avg_vol))

df_raw_vol$Region <- recode_factor(df_raw_vol$Region,
                           left_anterior_hippocampus= "L Ant Hip",
                           right_anterior_hippocampus = "R Ant Hip",
                           left_posterior_hippocampus = "L Post Hip",
                           right_posterior_hippocampus = "R Post Hip",
                           left_br36 = "L Br36",
                           right_br36 = "R Br36",
                           left_phc = "L PHC",
                           right_phc = "R PHC",
                           left_br35 = "L Br35",
                           right_br35 = "R Br35",
                           left_erc = "L ERC",
                           right_erc = "R ERC")

df_raw_vol$ad_present <- recode_factor(df_raw_vol$ad_present,
                               Normal = "Control",
                               `FALSE` = "LBD-AD",
                               `TRUE` = "LBD+AD")


plot <-
  ggplot(df_raw_vol, aes(x = ad_present, y = Volume)) +
  geom_boxplot() +
  facet_wrap(.~Region, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Subregion Volumes, Unadjusted (After Batch Correction)",
       y = "Volume (cm^3)",
       x = "Case Status") +
  font("y", size = 15) +
  font("x", size = 15) +
  theme(strip.text.x = element_text(size = 7)) +
  geom_jitter(alpha = 0.2, size =1)

plot
```

## Regression - Control vs Case
```{r regression}
df <- ds_covbat_adjusted
title <- c("Batch Corrected divided by ICV (Control = 0, Case = 1)")

# Define comparison, i.e. control = 0, case = 1 
df1 <- df %>%  select(-contains(c("misc", "meninges"))) %>%
  mutate(case_status = ifelse(diagnosis == "Normal", 0,
                              ifelse(diagnosis != "Normal", 1,
                                      "NA")))

# Position of variables in data frame to be used in for loop
# region raw volume
vol_start <- grep("left_anterior_hippocampus$", colnames(df1)) #column index to begin taking outcome vars
vol_end <- grep("right_phc$", colnames(df1)) #column index to end
vol_nvar <- vol_end - vol_start + 1

# shift within df
out_shift <- vol_start - 1

#For loop
dat <- df1
out_mod <- vector('list', length(vol_nvar))
i <- vol_start

for (i in vol_start:vol_end){
  raw_volume = colnames(dat)[i]

mod_1 <- lm(get(raw_volume) ~ case_status,
      na.action = na.exclude,
      data=dat)

out_mod[[i-out_shift]] <- mod_1
names(out_mod)[i-out_shift] <- paste0("model_", colnames(dat)[i])
}


library(broom)

out_mod_summary <- summary(out_mod[[1]])
out_mod_summary$coefficients %>% as_tibble()
z <- tidy(mod_1, conf.int =T)[2,]
z %>% mutate(region = names(out_mod[1]))

tidy(out_mod[[1]])

names(out_mod[1])

df <- NULL

for (i in 1:length(out_mod)) {
z <- tidy(out_mod[[i]], conf.int =T)[2,]
z <- z %>% mutate(region = names(out_mod[i]))
df <- rbind(df, z)
}

df1 <- df %>% filter(!str_detect(region, pattern = "_sul|_misc|_meninges")) %>%
  arrange(p.value) %>%
  mutate(region = str_remove(region, "model_"),
         region = str_replace(region, "_", " ")) %>%
  select(Region = "region", `Estimate` = "estimate", p = "p.value", 
         `S.E.` = "std.error", `2.5% CI` = "conf.low", `97.5% CI` = "conf.high")

library(kableExtra)
table <- knitr::kable(df1,
  caption = title)

kable_classic_2(table)

```

## Regression - Batch corrected and divided by ICV - AD+ vs AD - 
```{r regression using unadjusted vols, echo=TRUE}
df <- ds_covbat_adjusted
title <- c("Batch Corrected divided by ICV (AD- = 0, AD+ = 1)")

# Define comparison, i.e. control = 0, case = 1 
df1 <- df %>%  select(-contains(c("misc", "meninges"))) %>%
  filter(ad_present != "Control") %>% 
  mutate(case_status = ifelse(ad_present == "LBD-AD", 0,
                              ifelse(ad_present == "LBD+AD", 1,
                                      "NA")))

# Position of variables in data frame to be used in for loop
# region raw volume
vol_start <- grep("left_anterior_hippocampus$", colnames(df1)) #column index to begin taking outcome vars
vol_end <- grep("right_phc$", colnames(df1)) #column index to end
vol_nvar <- vol_end - vol_start + 1

# shift within df
out_shift <- vol_start - 1

#For loop
dat <- df1
out_mod <- vector('list', length(vol_nvar))
i <- vol_start

for (i in vol_start:vol_end){
  raw_volume = colnames(dat)[i]

mod_1 <- lm(get(raw_volume) ~ case_status,
      na.action = na.exclude,
      data=dat)

out_mod[[i-out_shift]] <- mod_1
names(out_mod)[i-out_shift] <- paste0("model_", colnames(dat)[i])
}


library(broom)

out_mod_summary <- summary(out_mod[[1]])
out_mod_summary$coefficients %>% as_tibble()
z <- tidy(mod_1, conf.int =T)[2,]
z %>% mutate(region = names(out_mod[1]))

tidy(out_mod[[1]])

names(out_mod[1])

df <- NULL

for (i in 1:length(out_mod)) {
z <- tidy(out_mod[[i]], conf.int =T)[2,]
z <- z %>% mutate(region = names(out_mod[i]))
df <- rbind(df, z)
}

df1 <- df %>% filter(!str_detect(region, pattern = "_sul|_misc|_meninges")) %>%
  arrange(p.value) %>%
  mutate(region = str_remove(region, "model_"),
         region = str_replace(region, "_", " ")) %>%
  select(Region = "region", `Estimate` = "estimate", p = "p.value")

library(kableExtra)
table <- knitr::kable(df1,
  caption = title)

kable_classic_2(table)

```

# left + right divided by ICV 
