---
title: "Verbal memory analysis"
author: "Jesse Cohen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(randomForest)
library(janitor)
library(rattle)
library(corrr)
library(lobstr)       # Inspect R data structures.
library(FSelector)    # Feature selection, information.gain().
library(stringi)      # String concat operator %s+%.
library(naniar)

rm(list = ls())
```

Initialize per template

# Data Import Template
```{r initialize}
# input
verbal_testing_data <- read_rds(here("objects/vlt_raw.RDS"))
ashs_clean <- read_rds("objects/ashs_clean.RDS")

# Initialise the dataset as per the template.
dsname <- "verbal_testing_data"
ds     <- get(dsname)

ds %>% sample_frac()

glimpse(ds)
```

Normalize variable names automatically
```{r}
# Capture the original variable names for use in plots.

vnames <- names(ds)

# Normalise the variable names.

ds %<>% clean_names(numerals="right")

# Confirm the results are as expected.

names(ds)

# Index the original variable names by the new names.

names(vnames) <- names(ds)

vnames

# Note the available variables.

vars <- names(ds) %T>% print()

```

```{r}
# Note the identifiers.

id <- c("inddid", "inddid_1", "test_date", "visit_id")

# Initialise ignored variables: identifiers.

ignore <- c(id)

```

Character variables 
```{r}
# Identify the character variables by index.

ds %>%
  sapply(is.character) %>%
  which() %T>%
  print() ->
chari

# Identify the character variables by name.

ds %>% 
  names() %>% 
  '['(chari) %T>% 
  print() ->
charc

# Observe the unique levels.

ds[charc] %>% sapply(unique)


# Ignore character variables

ignore <- union(ignore, charc)

```

Once we have identified all of the variables to ignore we remove them from our list of variables to use.
```{r}
# Check the number of variables currently.

length(vars)

# Remove the variables to ignore.

vars <- setdiff(vars, ignore)

# Confirm they are now ignored.

length(vars)
```

Numeric Variables
```{r}
ds[vars] %>%
  sapply(is.numeric) %>%
  which() %>%
  names %T>%
  print() ->
numi

ds[numi] %>% 
  summary()
```

# Data Exploration

A random sample of the dataset:
```{r}
ds %>% sample_frac()
```


# Verbal memory
Caveats with combining the two scores: as opposed to HVLT, PVLT looks at effect of distractors - after trial_ 5 adds a distractor list. Will z-score within each test within domains of learning (immediate recall), delayed recall (including "rapid forgetting index" of number on delayed recall:best immediate recall score), recognition (recognition discrimination index).
-Need to show consistent relationships between domains in HVLT vs PVLT in order to legitimize combining them 
- will use score closest in time to MRI that occurs after MRI, since in mediation analysis the cognitive test has to come after the MRI  

Steps:
1. find VLT closest to MRI date but occurring after
2. z-scores for each domain:
PVLT learning  - 1-5 trial_s total. also z-scores for each individual trial_ and then average those z scores
delayed - (Trial 9)
recognition - recognition discrimination score

HVLT learning - Total recall score. also z-scores for each individual trial_ then average those z scores
delayed - including Rapid forgetting index (%retained - measures what is lost between highest score in immediate recall and delayed recall)
recognition - recognition discrimination index 

--> subsequently can include MOCA/MMSE z-scores in the final model to adjust for global cognition


```{r verbal memory}
#find verbal memory test closest to mri date for each subject

verbal_date <- ashs_clean %>% 
  left_join(ds, by="inddid") %>%
  filter(diagnosis != "Normal") %>% 
  mutate(session_date = as.Date(session_date, format = '%Y%m%d'))

verbal_date %<>% 
  select(inddid, diagnosis, ad_present, session_date, hvlt_date = test_date, pvlt_date = `test_date_1`, dplyr::all_of(numi)) %>%
  filter(is.na(phl_recognition_discrimin) == F) %>% #remove individuals who couldn't complete the test
  pivot_longer(cols = c("hvlt_date", "pvlt_date"), names_to = "test", values_to = "test_date") %>% 
  mutate(intvl_test = as.numeric(as.Date(test_date) - as.Date(session_date))/365) %>% 
  select(inddid, diagnosis, ad_present, session_date, intvl_test, test, test_date) %>% distinct() %>% 
  filter(is.na(intvl_test)==F) %>% 
  filter(intvl_test > 0) %>% 
  group_by(inddid) %>% 
  filter(intvl_test == min(intvl_test, na.rm = T)) %>%
  ungroup() 

```


```{r}
#correlation of corrected volumes with HVLT scores
# for now with uncorrected volumes

# HVLT learning - Total recall score. also z-scores for each individual trial_ then average those z scores
# delayed - including Rapid forgetting index (%retained - measures what is lost between highest score in immediate recall and delayed recall)
# recognition - recognition discrimination index 

# Subset HVLT testing
hvlt <- ds %>% select(inddid, test_date, ageat_test, dplyr::all_of(dplyr::contains("hvlt"))) %>% 
  drop_na() %>% 
  distinct()
  
# First calculate z-scores for all cases (subsequently will look at just those closest to MRI)
# Select hvlt tests to use and add columns for max immediate recall score, "rapid forgetting index" (delayed recall score/max immediate recall score)
data <- hvlt %>% 
  select(inddid, test_date, hvltrt_1, hvltrt_2, hvltrt_3, hvltrt_tot_score, hvltrdly, hvltrecdist_score) %>% 
  rowwise() %>% 
  mutate(immediate_max = max(hvltrt_1, hvltrt_2, hvltrt_3)) %>% 
  mutate(remember_index = hvltrdly/immediate_max)

z_scores <- sapply(data[,-(1:2)], function(data) (data-mean(data, na.rm = T))/sd(data, na.rm = T))

#rename z_score column names 
original_cols <- colnames(z_scores) 
colnames(z_scores) <- paste0(original_cols, "_z_score")

data <- cbind(data, z_scores)
 
# #rejoin to original df
# hvlt_normed <- verbal_date %>% filter(test == "hvlt_date") %>% left_join(., data) %>% as_tibble()

#create new column for mean of immediate recall z-scores
hvlt_normed <- data %>% 
  rowwise() %>% 
  mutate(immediate_z_score_mean = mean(hvltrt_1_z_score, hvltrt_2_z_score, hvltrt_3_z_score, na.rm = T)) 
  
### Library
library(heatmaply)

### Data
df <- hvlt_normed %>% select(contains("z_score"))

### Let's Plot
hvlt_heatmap <- heatmaply_cor(x = cor(df),
              xlab = "Features",
              ylab = "Features",
              k_col = 2,
              k_row = 2,
              main = "HVLT Z-scores")

hvlt_heatmap
```

```{r}
# PVLT
# Subset PVLT testing
pvlt <- ds %>% select(inddid, test_date_1, ageat_test, dplyr::all_of(dplyr::contains("phl"))) %>% 
  replace_with_na(replace = list("<NA>")) %>% 
  filter(!is.na(test_date_1)) %>% 
  distinct()

# #remove individuals who couldn't complete the test
# pvlt <- pvlt %>% filter(is.na(phl_recognition_discrimin)==F)

data <- pvlt %>% 
  select(inddid, test_date_1, phl_correct_trial_1, phl_correct_trial_2, phl_correct_trial_3, phl_correct_trial_4, phl_correct_trial_5,
         phl_correct_15total, phl_correct_trial_9, phl_recognition_discrimin) %>% 
  filter(is.na(phl_recognition_discrimin)==F) %>% 
  rowwise() %>% 
  mutate(immediate_max = max(phl_correct_trial_1, phl_correct_trial_2, phl_correct_trial_3, phl_correct_trial_4, phl_correct_trial_5)) %>% 
  mutate(remember_index = phl_correct_trial_9/immediate_max) 

z_scores <- sapply(data[,-(1:2)], function(data) (data-mean(data))/sd(data))
  
#rename z_score column names 
original_cols <- colnames(z_scores) 
colnames(z_scores) <- paste0(original_cols, "_z_score")

data <- cbind(data, z_scores)
# 
# #rejoin to original df
# pvlt_normed <- verbal_date %>% filter(test == "pvlt_date") %>%
#   left_join(., data) %>% as.tibble()

#create new column for mean of immediate recall z-scores
pvlt_normed <- data %>% 
  rowwise() %>% 
  mutate(immediate_z_score_mean = mean(phl_correct_trial_1_z_score, phl_correct_trial_2_z_score, phl_correct_trial_3_z_score, phl_correct_trial_4_z_score, phl_correct_trial_5_z_score)) 

### Data
df <- pvlt_normed %>% select(contains("z_score"))

### Let's Plot
pvlt_heatmap <- heatmaply_cor(x = cor(df),
              xlab = "Features",
              ylab = "Features",
              k_col = 3,
              k_row = 3,
              main = "PVLT Z-scores")


hvlt_heatmap
pvlt_heatmap

# Create a heatmap of individuals with both pvlt and hvlt.
# Select a single score for total, delay, memory, and recall
ds_h <- hvlt_normed %>% select(inddid, test_date,
                       h_1 = hvltrt_1_z_score, 
                       h_total = hvltrt_tot_score_z_score,
                       h_recall = remember_index_z_score,
                       h_recog = hvltrecdist_score_z_score)

ds_p <- pvlt_normed %>% select(inddid, test_date_1,
                               p_1 = phl_correct_trial_1_z_score,
                               p_total = phl_correct_15total_z_score,
                               p_recall = remember_index_z_score,
                               p_recog = phl_recognition_discrimin_z_score)

# Select ids with both tests completed at least once
both_vlt <- intersect(hvlt_normed$inddid, pvlt_normed$inddid)

ds_z <- 
  ds_h %>% full_join(ds_p) %>% 
  filter(inddid %in% both_vlt) %>% 
  mutate(date_diff = abs(as.numeric(test_date - test_date_1)/365)) 

df <- ds_z %>% 
  group_by(inddid) %>% 
  filter(date_diff == min(date_diff)) %>% 
  ungroup(inddid) %>% 
  select(-c(inddid, test_date, test_date_1, date_diff))


### Let's Plot
combo_heatmap <- heatmaply_cor(x = cor(df),
              xlab = "Features",
              ylab = "Features",
              k_col = 3,
              k_row = 3,
              main = "VLT Z-scores")

combo_heatmap
```

# correlation of PVLT with volumes
```{r correlation of VLT with subfield volumes}

pvlt_ashs <-
  verbal_date %>% 
  filter(test == "pvlt_date") %>% 
  left_join(ds_p, by = c(inddid = "inddid", test_date = "test_date_1")) %>% 
  distinct() %>% 
  left_join(ashs_clean)

ignore <- c("test_date", "yob", "intvl_test", "session_year")

pvlt_ashs %<>% 
  select(-contains(c("misc", "sul", "meninges")),
         -any_of(ignore))

vnames <- quo(c("L Ant Hip" = left_anterior_hippocampus,
"R Ant Hip" = right_anterior_hippocampus,
"L Post Hip" = left_posterior_hippocampus,
"R Post Hip" = right_posterior_hippocampus,
"L Br36" = left_br_36,
"R Br36" = right_br_36,
"L PHC" = left_phc,
"R PHC" = right_phc,
"L Br35" = left_br_35,
"R Br35" = right_br_35,
"L ERC" = left_erc,
"R ERC" = right_erc,
"ICV" = left_icv))

pvlt_ashs %<>% rename(!!vnames)

p <- pvlt_ashs %>%
  correlate() %>%
  rearrange() %>%
  shave() %>% 
  rplot() 
  
p + guides(x = guide_axis(n.dodge = 2))

pvlt_ashs %>%
  correlate() %>%
  rearrange() %>% 
  focus(contains("p_")) 

pvlt_ashs %>%
  corrr::correlate() %>%
  shave() %>%
  stretch() %>%
  filter(abs(r) > 0.60) %>% 
  arrange(desc(r))

pvlt_ashs %>%
  correlate() %>%
  network_plot()


# Correlation plots

p_long <- pvlt_ashs %>% 
  pivot_longer(cols = contains("p_"), names_to = "test_component",
                           values_to = "z_score") %>% 
  pivot_longer(cols = contains(c("L ", "R ", "ICV")),
               names_to = "region",
               values_to = "vol") %>% 
  filter(test_component != "p_total") #remove p_total 

p_hip <- p_long %>% filter(stringr::str_detect(region, "Hip"))
p_br36 <- p_long %>% filter(stringr::str_detect(region, "Br36"))
p_br35 <- p_long %>% filter(stringr::str_detect(region, "Br35"))
p_phc <- p_long %>% filter(stringr::str_detect(region, "PHC"))

vol_vlt_cor <- function(long_df) {

ggscatter(long_df, x = "vol", y = "z_score",
          facet.by = c("region", "test_component"),
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Volume", 
          ylab = "PVLT Z-score", 
          title = "Memory and MTL Subfields")
   
}

vol_vlt_cor(p_hip)
vol_vlt_cor(p_br36)
vol_vlt_cor(p_br35)
vol_vlt_cor(p_phc)


```

# Correlation with HVLT with volumes
```{r}
hvlt_ashs <-
  verbal_date %>% 
  filter(test == "hvlt_date") %>% 
  left_join(ds_h, by = c(inddid = "inddid", test_date = "test_date")) %>% 
  distinct() %>% 
  left_join(ashs_clean)

ignore <- c("test_date", "yob", "intvl_test", "session_year")

hvlt_ashs %<>% 
  select(-contains(c("misc", "sul", "meninges")),
         -any_of(ignore))

vnames <- quo(c("L Ant Hip" = left_anterior_hippocampus,
"R Ant Hip" = right_anterior_hippocampus,
"L Post Hip" = left_posterior_hippocampus,
"R Post Hip" = right_posterior_hippocampus,
"L Br36" = left_br_36,
"R Br36" = right_br_36,
"L PHC" = left_phc,
"R PHC" = right_phc,
"L Br35" = left_br_35,
"R Br35" = right_br_35,
"L ERC" = left_erc,
"R ERC" = right_erc,
"ICV" = left_icv))

hvlt_ashs %<>% rename(!!vnames)

p <- hvlt_ashs %>%
  correlate() %>%
  rearrange() %>%
  shave() %>% 
  rplot() 
  
p + guides(x = guide_axis(n.dodge = 2))

hvlt_ashs %>%
  correlate() %>%
  rearrange() %>% 
  focus(contains("p_")) 

hvlt_ashs %>%
  corrr::correlate() %>%
  shave() %>%
  stretch() %>%
  filter(abs(r) > 0.60) %>% 
  arrange(desc(r))

hvlt_ashs %>%
  correlate() %>%
  network_plot()


# Correlation plots

h_long <- hvlt_ashs %>% 
  pivot_longer(cols = contains("h_"), names_to = "test_component",
                           values_to = "z_score") %>% 
  pivot_longer(cols = contains(c("L ", "R ", "ICV")),
               names_to = "region",
               values_to = "vol") %>%  
  filter(test_component != "h_total") #remove p_total 
  

h_hip <- h_long %>% filter(stringr::str_detect(region, "Hip"))
h_br36 <- h_long %>% filter(stringr::str_detect(region, "Br36"))
h_br35 <- h_long %>% filter(stringr::str_detect(region, "Br35"))
h_phc <- h_long %>% filter(stringr::str_detect(region, "PHC"))

# Reuse function created above
vol_vlt_cor(h_hip)
vol_vlt_cor(h_br36)
vol_vlt_cor(h_br35)
vol_vlt_cor(h_phc)
```

